Renaming The Amplicons and Tidying up the Amplicon Information Table

### 2024-06-18 extras/Renaming-amplicons/Amplicon-renaming.Rmd
Done with revamping the locus names

This is an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code.

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*.

```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

>>>>

Before releasing this, we want to have a decent, consistent naming convention for
all of the amplicons.  We want a short, unique, descriptive name that will tell us
what kind of marker it is (snplicon vs microhap, etc), and then we want to have
position information in there too.

Here is the convention that we came up with:

General pattern = `Ots_type00X_ChrNum_V2StartPos`, where:

- `Ots` is just `Ots`
- `type` is a brief descriptor (shoot for 4 characters) for example:
    + SNPtype = `scon`
    + Microhap = `mhap`
    + RoSA = `rosa`
    + SexID = `sexy`
    + LFAR = `lfar`
    + WRAP = `wrap`
    + VGLL = `vgll`
    + Six6 = `sixx`
- 00X is a three digit left zero-padded, base-1 number within type ordered by genome V2 location.
- ChrNum is a two digit left zero padded Chromosome number, except for sexy which is
  going to be just the scaffold name, with the underscore removed.
- V2StartPos is just where it maps in Otsh_V2.

So, let's get crackin' on slurping stuff out of the Excel file and making these modifications.

## Doing it

```{r}
library(tidyverse)
library(readxl)

# read it in and change some of the names
orig <- read_excel("ChinookFullPanelLociTable_May2024Updated.xlsx") %>%
  rename(
    OtherName = Locus,
    Otsh_v1.0_coordinates = AmpliconGenomeV1Location,
    Otsh_v2.0_coordinates = AmpliconGenomveV2Location,
    Alternate_Otsh_v1.0_coordinates = `AlternateGenomeV1Location(s)`,
    Alternate_Otsh_v2.0_coordinates = `AlternateGenomeV2Location(s)`,
    LocusOrigin = LocusType,
    OriginalReference = `Original Reference`,
    MappingNotes = Notes
  )

# now, reorder the columns so that more information is front loaded
orig2 <- orig %>%
  select(
    LocusOrigin,
    OriginalReference,
    OtherName,
    ForwardPrimerSequence,
    ReversePrimerSequence,
    ReferenceSequence,
    SNPLocationsInReferenceSequence,
    TrimmedAmpliconSequence,
    Otsh_v2.0_coordinates,
    Alternate_Otsh_v2.0_coordinates,
    Otsh_v1.0_coordinates,
    Alternate_Otsh_v1.0_coordinates,
    everything()
  )
```

Now we are going to get columns of chromosome number (or name if unmapped) and
start positions, so that we can order things by genome coordinate.

```{r}
# get the chromosome numbers (pulled from NCBI page at https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_018296145.1/)
chroms <- read_tsv("Otsh_v2.0_chroms.tsv") %>%
  mutate(Otsh_v2.0_chromosome = str_replace(Chromosome, "LG", ""), .before = Chromosome) %>%
  select(RefSeq, Otsh_v2.0_chromosome)

orig3 <- orig2 %>%
  extract(Otsh_v2.0_coordinates, into = c("RefSeq", "AmpliconStartPos"), regex = "^(.*):([0-9]+)-[0-9]+$", remove = FALSE, convert = TRUE) %>%
  left_join(chroms, by = join_by(RefSeq)) %>%
  mutate(Otsh_v2.0_chromosome = ifelse(is.na(Otsh_v2.0_chromosome), str_replace(RefSeq, "_", ""), Otsh_v2.0_chromosome)) %>%
  select(Otsh_v2.0_chromosome, AmpliconStartPos, everything()) %>%
  select(-RefSeq) %>%
  arrange(Otsh_v2.0_chromosome, AmpliconStartPos)

```

Now we
will be able to rename everything all at once, easily.

```{r}
final <- orig3 %>%
  mutate(
    tmpname = case_match(
      LocusOrigin,
      "LFAR" ~ "lfar",
      "Microhap" ~ "mhap",
      "RoSA" ~ "rosa",
      "SNPtype" ~ "scon",
      "Sex ID" ~ "sexy",
      "Six6" ~ "sixx",
      "VGLL3" ~ "vgll",
      "WRAP" ~ "wrap"
    )
  ) %>% 
  group_by(tmpname) %>%
  mutate(
    AmpliconName = str_c(
      "Ots_", tmpname,
      sprintf("%03d", 1:n()), "_", 
      Otsh_v2.0_chromosome, "_", 
      sprintf("%08d",AmpliconStartPos)),
    .before = Otsh_v2.0_chromosome
  ) %>%
  ungroup() %>%
  select(-tmpname)
```

All right!  That pretty much does it.  Let's just print it out.

```{r}
write_csv(final, "../../inputs/Calif-Chinook-Amplicon-Panel-Information.csv", na = "")
```
